---
title: "Avalanche Project"
output: word_document
date: "2026-02-04"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Goal: Data analysis project using Bayesian statistics on Colorado Avalanche data.

# Background 
```{r}
#load libraries 
library(tidyverse)
library(ggplot2)
library(arrow) # to read in data 
library(dplyr)
library(brms) #bayesian regression
library(tidybayes) #posterior
library(loo) #model comparison 
library(patchwork) #combing plots 
library(ggrepel) #non-overlapping labels on quadrant plots
library(knitr) # for nice table 
```

```{r}
#load data
players <- read_parquet("~/personal/NHL/players.parquet") #one record per game with team and outcome information
games <-  read_parquet("~/personal/NHL/games.parquet") #one record per game with team and outcome information
stints <-  read_parquet("~/personal/NHL/stints.parquet") #one record per player per stint (a stretch of play during which a set of distinct players is on the ice)
events <-  read_parquet("~/personal/NHL/events.parquet") #one record per event
tracking <-  read_parquet("~/personal/NHL/tracking.parquet") #one record per player per event
```

*Datasets Info.*
Players: 1172 obs. of 8 variables 
player_id, player_name, last_name, first_name, handed, birth_date, position_group, primary_position 

Games: 480 obs. of 12 variables 
game_id, game_date, league, season, home_team, away_team, home_team_id, away_team_id, home_score, away_score, game_outcome, home_start_net 

Stints: 480 obs. of 12 variables
game_id, game_date, league, season, home_team, away_team, home_team_id, away_team_id, home_score, away_score, game_outcome, home_start_net 

Events: 1800464 obs. of 24 variables 
game_id, period, period_time, game_stint, sl_event_id, sequence_id, player_id, player_name, team, team_id, opp_team, opp_team_id, event_type, outcome, flags, description, detail, sl_xg_all_shots, x, y, x_adj, y_adj, has_tracking_data, event_player_tracked   

Tracking: 13529224 obs. of 10 variables
game_id, sl_event_id, team_id, team_name, player_id, player_name, tracking_x, tracking_y, tracking_vel_x, tracking_vel_y 


##1st  How do expected goals (xG) very by location and player?

# Part 1 estimated a static xG surface: how shot location drives goal probability,
# with player-level random intercepts capturing individual deviations from that surface.
Model: Bayesian hierarchical logistic regression 
-modeling shot success probability as a function of spatial stats w/ partial pooling by player

# Preprocessing Data 
```{r}
#filter + prepare shot data 
shots <- events %>%
  filter(
    event_type %in% c("shot", "goal"),
    !is.na(x_adj),
    !is.na(y_adj),
    !is.na(outcome),
    !is.na(player_id)
  ) %>%
  mutate(
    goal = as.integer(event_type == "goal"),  #target: goal scored or not
    x_scaled = scale(x_adj)[,1],
    y_scaled = scale(y_adj)[,1]
  )
##RERUN FOR THIS 
shots <- shots %>%
  mutate(
    # Flip coordinates so all shots attack in the same direction
    x_offense = abs(x_adj),        # distance from own goal always positive
    y_offense = ifelse(x_adj > 0, -y_adj, y_adj),  # mirror y accordingly
    
    #rescale 
    x_scaled = scale(x_offense)[,1],
    y_scaled = scale(y_offense)[,1]
  )
```

```{r}
count(shots, event_type, goal)
shots %>% count(goal) %>% mutate(pct = n / sum(n)) # making sure distribution is correct % of goals scored to not 
```


```{r}
#model formula
shot_formula <- bf(
  goal ~ x_scaled + y_scaled + I(x_scaled^2) + I(y_scaled^2) + (1| player_id) #random intercept
)
```

Fixed effects capture the population-level shooting probability as a function of location. 
The quadratic terms let the xG surface curve.

(1 | player_id) is the partial pooling term
-each player gets their own intercept offset

```{r eval = FALSE}
#set priors 
priors <- c(
  prior(normal(0, 1.5), class = b),           # fixed effect coeff.
  prior(normal(-2, 1),  class = Intercept),   # baseline log-odds; ~12% base rate
  prior(exponential(1), class = sd)           # SD player random effects
)
```

For prior, league-wide shooting % is roughly 10–12%, which on the log-odds scale is about log(0.11 / 0.89) ≈ -2.1. 
-rules out crazy values

https://cran.r-project.org/bin/windows/Rtools/

```{r eval = FALSE}
#Save and reload the compiled fit object
if (!file.exists("fit.rds")) {
  fit <- brm(
  formula  = shot_formula,
  data     = shots,
  family   = bernoulli(link = "logit"),
  prior    = priors,
  chains   = 2,
  iter     = 1000,          # 500 warmup + 500 sampling per chain
  cores    = 4,   # for speed 
  seed     = 42,
  control  = list(adapt_delta = 0.95)   # reduce divergences on hierarchical models
)
  saveRDS(fit, "fit.rds")
} else {
  fit <- readRDS("fit.rds")
}
```

```{r eval = FALSE}
#check the model
summary(fit)  #look at Rhat 

#Posterior predictive check
pp_check(fit, ndraws = 20)
```

```{r eval = FALSE}
# compare to recorded xG
# Check the variable exists and its range
shots %>%
  filter(!is.na(sl_xg_all_shots)) %>%
  summarise(
    n        = n(),
    mean_xg  = mean(sl_xg_all_shots),
    min_xg   = min(sl_xg_all_shots),
    max_xg   = max(sl_xg_all_shots)
  )

#extract models predictions + join them
#posterior mean xG for every shot
shots$xg_model <- posterior_epred(
  fit,
  newdata        = shots,
  re_formula     = NA,          # population-level only
  allow_new_levels = TRUE
) %>% colMeans()

# Join + compare
comparison <- shots %>%
  filter(!is.na(sl_xg_all_shots)) %>%
  select(player_id, player_name, goal, xg_model, sl_xg_all_shots)
```

```{r eval = FALSE}
#run comparisons 
#Correlation
cor(comparison$xg_model, comparison$sl_xg_all_shots, use = "complete.obs")

#plot
ggplot(comparison, aes(x = sl_xg_all_shots, y = xg_model)) +
  geom_point(alpha = 0.1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "Model xG vs Avs xG",
    x     = "Avs xG ",
    y     = "Bayesian xG"
  )

#calibration
comparison %>%
  summarise(
    your_xg_mean      = mean(xg_model),
    avs_xg_mean       = mean(sl_xg_all_shots)
  )
```

```{r eval = FALSE}
# visual of xG surface 
#grid over the rink
grid <- expand_grid(
  x_offense = seq(min(shots$x_offense), max(shots$x_offense), length.out = 80),
  y_offense = seq(min(shots$y_offense), max(shots$y_offense), length.out = 80)
) %>%
  mutate(
    x_scaled = (x_offense - mean(shots$x_offense)) / sd(shots$x_offense),
    y_scaled = (y_offense - mean(shots$y_offense)) / sd(shots$y_offense),
    player_id = NA
  )

grid$xg <- posterior_epred(fit, newdata = grid, re_formula = NA, allow_new_levels = TRUE) %>%
  colMeans()

#plot using x_offense/y_offense now
ggplot(grid, aes(x_offense, y_offense, fill = xg)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", labels = scales::percent) +
  labs(title = "Posterior Mean xG Surface", fill = "xG", 
       x = "x (offense)", y = "y (offense)")
```

```{r eval = FALSE}
#Player-level random effects
player_effects <- fit %>%
  spread_draws(r_player_id[player_id, ])  %>%
  median_qi(.width = 0.95) %>%
  left_join(
    shots %>% distinct(player_id, player_name) %>% count(player_id, player_name),
    by = "player_id"
  ) %>%
  arrange(desc(r_player_id))

#Top 15 overperformers vs bottom 15
player_effects %>%
  slice(c(1:15, (n()-14):n())) %>%
  ggplot(aes(x = r_player_id, y = reorder(player_name, r_player_id),
             xmin = .lower, xmax = .upper)) +
  geom_pointrange() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Player random effects: shooting skill beyond location",
       x = "Random intercept (log-odds)", y = NULL)
```

## TRY AGAIN : ADJUSTED PRIOR
*same just new prior*
```{r}
#TRY AGAIN new PRIOR
#set priors 
priors2 <- c(
  prior(normal(0, 1.5), class = b),           # fixed effect coeff.
  prior(normal(-2, 1),  class = Intercept),   # baseline log-odds; ~12% base rate
  prior(exponential(2), class = sd)           # SD player random effects  ##CHANGE to exponential(2)
)

#Save + reload 
if (!file.exists("new_fit.rds")) {
  new_fit <- brm(
  formula  = shot_formula,
  data     = shots,
  family   = bernoulli(link = "logit"),
  prior    = priors2,
  chains   = 2,
  iter     = 1000,          # 500 warmup + 500 sampling per chain
  cores    = 4,   # for speed 
  seed     = 42,
  control  = list(adapt_delta = 0.95)   # reduce divergences on hierarchical models
)
  saveRDS(fit, "new_fit.rds")
} else {
  new_fit <- readRDS("new_fit.rds")
}
```

```{r}
summary(new_fit)
plot(new_fit)
```

```{r}
#Posterior predictive check
pp_plot <- pp_check(new_fit, ndraws = 100)
print(pp_plot)
```

Compare to recorded xG.
```{r}
# Check the variable exists + range
shots %>%
  filter(!is.na(sl_xg_all_shots)) %>%
  summarise(
    n        = n(),
    mean_xg  = mean(sl_xg_all_shots),
    min_xg   = min(sl_xg_all_shots),
    max_xg   = max(sl_xg_all_shots)
  )

#extract models predictions + join 
#posterior mean xG for every shot
shots$xg_model <- posterior_epred(
  new_fit,
  newdata        = shots,
  re_formula     = NA,          # population-level only
  allow_new_levels = TRUE
) %>% colMeans()

# Join + compare
comparison <- shots %>%
  filter(!is.na(sl_xg_all_shots)) %>%
  select(player_id, player_name, goal, xg_model, sl_xg_all_shots)
```

Run comparisons
```{r}
#correlation
cor(comparison$xg_model, comparison$sl_xg_all_shots, use = "complete.obs")

#plot
comp_plot <- ggplot(comparison, aes(x = sl_xg_all_shots, y = xg_model)) +
  geom_point(alpha = 0.1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "xG Bayesian Model vs Avs xG",
    x     = "Avs xG",
    y     = "Bayesian xG"
  )
print(comp_plot)

#calibration
comparison %>%
  summarise(
    my_xg_mean      = mean(xg_model),
    avs_xg_mean       = mean(sl_xg_all_shots)
  )
```

Visual heatmap of xG surface.
```{r}
#grid over the rink
grid <- expand_grid(
  x_adj = seq(min(shots$x_adj), max(shots$x_adj), length.out = 80),
  y_adj = seq(min(shots$y_adj), max(shots$y_adj), length.out = 80)
) %>%
  mutate(
    x_scaled = (x_adj - mean(shots$x_adj)) / sd(shots$x_adj),
    y_scaled = (y_adj - mean(shots$y_adj)) / sd(shots$y_adj),
    player_id = NA    # population-level prediction 
  )

#Posterior mean xG per grid cell
grid$xg <- posterior_epred(new_fit, newdata = grid, re_formula = NA, allow_new_levels = TRUE) %>%
  colMeans()

heatmap <- ggplot(grid, aes(x_adj, y_adj, fill = xg)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", labels = scales::percent) +
  labs(title = "Posterior Mean xG Surface", fill = "xG", x = "x (adjusted)", y = "y (adjusted)")
plot(heatmap)

##NEW
grid <- expand_grid(
  x_offense = seq(min(shots$x_offense), max(shots$x_offense), length.out = 80),
  y_offense = seq(min(shots$y_offense), max(shots$y_offense), length.out = 80)
) %>%
  mutate(
    x_scaled = (x_offense - mean(shots$x_offense)) / sd(shots$x_offense),
    y_scaled = (y_offense - mean(shots$y_offense)) / sd(shots$y_offense),
    player_id = NA
  )

grid$xg <- posterior_epred(new_fit, newdata = grid, re_formula = NA, allow_new_levels = TRUE) %>%
  colMeans()

#Plot using x_offense/y_offense now
ggplot(grid, aes(x_offense, y_offense, fill = xg)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", labels = scales::percent) +
  labs(title = "Posterior Mean xG Surface", fill = "xG", 
       x = "x (offense)", y = "y (offense)")
```

Uncertainity heatmap visual.
```{r}
#full posterior draws per grid cell
pred_draws <- posterior_epred(
  new_fit, 
  newdata = grid, 
  re_formula = NA, 
  allow_new_levels = TRUE
)

#Mean surface
grid$xg_mean <- colMeans(pred_draws)

#Uncertainty surface
grid$xg_sd <- apply(pred_draws, 2, sd)

#uncertainty
uncertain_heatmap <- ggplot(grid, aes(x_offense, y_offense, fill = xg_sd)) +
  geom_tile() +
  scale_fill_viridis_c(option = "mako", labels = scales::percent) +
  labs(
    title = "Posterior Uncertainty: xG Surface SD",
    fill  = "SD",
    x     = "x (offense)",
    y     = "y (offense)"
  )
plot(uncertain_heatmap)
```

```{r}
#Player-level random effects
player_effects <- new_fit %>%
  spread_draws(r_player_id[player_id, ])  %>%
  median_qi(.width = 0.95) %>%
  left_join(
    shots %>% distinct(player_id, player_name) %>% count(player_id, player_name),
    by = "player_id"
  ) %>%
  arrange(desc(r_player_id))

# Top 15 overperformers vs bottom 15
player_effects %>%
  slice(c(1:15, (n()-14):n())) %>%
ggplot(aes(x = r_player_id, y = reorder(player_name, r_player_id),
             xmin = .lower, xmax = .upper)) +
  geom_pointrange() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Player Random Effects: shooting skill beyond location", 
       x = "Random intercept (log-odds)", y = NULL)
print(player_effects)
```




## Deeper Research Question
# Part 2 asks: is that surface stable across a season?
# Specifically, does a player's shooting probability from a given location
# drift across 10-game windows, or is the part 1 snapshot reliable?

# The key estimand is sigma_drift: the SD of how much each player's xG-location relationship shifts over time. 

Build longitudinal shot data. 
```{r}
#assign each shot to a time block (10-game windows)
#create a chronological game index per player
shots_long <- shots %>%
  # bring in game_date from games table
  left_join(games %>% select(game_id, game_date), by = "game_id") %>%
  arrange(player_id, game_date) %>%
  group_by(player_id) %>%
  mutate(
    # rank each game chronologically per player (1st game, 2nd game)
    game_rank  = dense_rank(game_date),
    # assign to 10-game windows: games 1-10 = block 1, 11-20 = block 2, etc
    time_block = ceiling(game_rank / 10)
  ) %>%
  ungroup() %>%
  group_by(player_id) %>%
  # only keep players with at least 2 time blocks
  filter(n_distinct(time_block) >= 2) %>%
  ungroup() %>%
   # create a clean player index 
  mutate(
    player_idx = as.integer(factor(player_id)),
    
    # USE x_offense/y_offense
    x_scaled = (x_offense - mean(shots$x_offense)) / sd(shots$x_offense),
    y_scaled = (y_offense - mean(shots$y_offense)) / sd(shots$y_offense)
  )

# check data 
shots_long %>%
  group_by(player_id) %>%
  summarise(
    n_shots  = n(),
    n_blocks = n_distinct(time_block),
    n_goals  = sum(goal)
  ) %>%
  summary()
```

Aggregate to player block level. 
```{r}
#model goal rate per block rather than individual shots
player_blocks <- shots_long %>%
  group_by(player_id, player_name, time_block) %>%
  summarise(
    goals      = sum(goal),         # goals scored in block
    shots      = n(),               # shots taken in block
    # average shot location per block (controls for quality of chances)
    mean_x     = mean(x_scaled),
    mean_y     = mean(y_scaled),
    .groups    = "drop"
  ) %>%
  # create a lagged block for the AR(1) structure, the previous block's index
  group_by(player_id) %>%
  arrange(time_block) %>%
  mutate(
    prev_block = lag(time_block),   # NA for first block of each player
    block_seq  = row_number()       # sequential block # within player
  ) %>%
  ungroup()

#structure
glimpse(player_blocks)
player_blocks %>%
  summarise(
    mean_shots = mean(shots),
    mean_goals = mean(goals),
    min_shots  = min(shots),
    max_shots  = max(shots)
  )
```
```{r}
player_blocks %>%
  count(shots < 5)
```

Define dynamic model
```{r}
dynamic_formula <- bf(
  goals | trials(shots) ~
    mean_x + mean_y +
    (time_block || player_id)) #removes the intercept-slope correlation since it was unidentifiable
```

Set priors for dynamic model on sigma_drift prior 
```{r}
dynamic_priors <- c(
  # ~5% shooting rate baseline on log-odds scale, consistent w/ part 1
  prior(normal(-3, 1),    class = Intercept),  #bc block-level binomial model
  #fixed effects for shot location same as pt 1 
  prior(normal(0, 1),     class = b),

  # between-player baseline differences, consistent with part 1
  prior(exponential(2),   class = sd, group= player_id, coef=Intercept),
  
  # sigma_drift do not think it will be large 
  #exponential(2) keeps prior mass near zero while allowing real signal
  prior(exponential(2), class=sd, group=player_id, coef= time_block)
)
```

Fit dynamic model.
```{r}
#Save + reload 
if (!file.exists("fit_dynamicmod.rds")) {
  fit_dynamicmod <- brm(
  formula   = dynamic_formula,
  data      = player_blocks,
  family    = binomial(link = "logit"),  # goals out of shots trials
  prior     = dynamic_priors,
  chains    = 4,
  iter      = 3000,         # more iterations than Q1 
  warmup    = 1500,
  cores     = 4,
  seed      = 42,
  control   = list(
    adapt_delta   = 0.97,   # higher than Q1 b/c hierarchical slopes need more tuning
    max_treedepth = 12
  )
)
  saveRDS(fit_dynamicmod, "fit_dynamicmod.rds")
} else {
  fit_dynamicmod <- readRDS("fit_dynamicmod.rds")
}  #toke 6 minutes 
```

```{r}
#diagnostics 
summary(fit_dynamicmod)  # Rhat
plot(fit_dynamicmod, ask = FALSE)  #Trace plots 

# Posterior predictive check
pp_check(fit_dynamicmod, ndraws = 100) +
  labs(title = "Posterior predictive check — dynamic model")
```
sigma_drift = 0.03 with 95% CI (0.00, 0.07) 
The Part 1 static xG surface is confirmed as a reliable seasonal picture. Player shooting profiles do not meaningfully drift across 10-game windows.

```{r}
player_blocks %>% count(goals) %>% arrange(goals)
```

#Answer question
```{r}
##how much does a player's xG surface shift across 10-game windows beyond what location alone predicts?

sigma_drift <- as_draws_df(fit_dynamicmod) %>%
  select(sigma_drift = `sd_player_id__time_block`)

sigma_plot <- ggplot(sigma_drift, aes(x = sigma_drift)) +
  stat_halfeye(.width = c(0.66, 0.95), fill = "#800000") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title    = "Posterior Distribution of xG Surface Drift (σ_Drift)",
    x        = "σ_Drift (log-odds scale)",
    y        = NULL
  )
print(sigma_plot)
#numerical summary
sigma_drift %>%
  summarise(
    median = median(sigma_drift),
    lower  = quantile(sigma_drift, 0.025),
    upper  = quantile(sigma_drift, 0.975)
  )
```

```{r eval = FALSE}
player_trajectories <- fit_dynamicmod %>%
  spread_draws(r_player_id[player_id, term]) %>%
  filter(term %in% c("Intercept", "time_block")) %>%
  pivot_wider(names_from = term, values_from = r_player_id) %>%
  left_join(
    player_blocks %>% distinct(player_id, player_name, time_block),
    by = "player_id"
  ) %>%
  mutate(
    skill_at_block = Intercept + time_block.y * time_block.x
  ) %>%
  group_by(player_id, player_name, time_block.y) %>%
  median_qi(skill_at_block, .width = 0.95) %>%
  rename(time_block = time_block.y)

# Top 12 players by shot volume
top_players <- player_blocks %>%
  group_by(player_id, player_name) %>%
  summarise(total_shots = sum(shots), .groups = "drop") %>%
  slice_max(total_shots, n = 12)

player_trajectories %>%
  filter(player_id %in% top_players$player_id) %>%
  ggplot(aes(x = time_block, y = skill_at_block)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.2, fill = "#5DCAA5") +
  geom_line(color = "#0F6E56", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ player_name, scales = "free_y") +
  labs(
    title    = "Player xG surface trajectories across the season",
    subtitle = "0 = population xG baseline for their shot locations\nShaded = 95% credible interval; flat line = stable spatial shooting profile",
    x        = "Time block (10-game windows)",
    y        = "Deviation from population xG surface (log-odds)"
  )
```

```{r eval = FALSE}
player_summary %>%
  arrange(desc(drift_median)) %>%
  slice(1:15) %>%
  ggplot(aes(x = drift_median, 
             y = reorder(player_name, drift_median),
             color = position_group)) +
  geom_pointrange(aes(
    xmin = quantile(drift_median, 0.025),
    xmax = quantile(drift_median, 0.975)
  )) +
  geom_vline(xintercept = median(player_summary$drift_median),
             linetype = "dashed") +
  labs(title = "Top 15 players by xG surface drift",
       x = "Drift magnitude (log-odds)", y = NULL)
```


```{r}
# Add shot volume +  CI width to player_summary
player_summary <- fit_dynamicmod %>%
  spread_draws(r_player_id[player_id, term]) %>%
  filter(term %in% c("Intercept", "time_block")) %>%
  pivot_wider(names_from = term, values_from = r_player_id) %>%
  group_by(player_id) %>%
  summarise(
    skill_median = median(Intercept),
    drift_median = median(abs(time_block)),
    skill_lower  = quantile(Intercept, 0.025),
    skill_upper  = quantile(Intercept, 0.975),
    .groups      = "drop"
  ) %>%
  left_join(
    players %>% select(player_id, player_name, position_group),
    by = "player_id"
  ) %>%
  # join shot volume for sizing
  left_join(
    shots_long %>% count(player_id, name = "n_shots"),
    by = "player_id"
  ) %>%
  #compute CI width as uncertainty measure
  mutate(ci_width = skill_upper - skill_lower)

# Store reference lines for quadrant colors
vline <- 0
hline <- median(player_summary$drift_median)

ggplot(player_summary,
       aes(x = skill_median, y = drift_median, color = position_group)) +
  
  #quadrant shading
  annotate("rect",
           xmin = 0,    xmax = Inf,
           ymin = -Inf, ymax = hline,
           fill = "green", alpha = 0.06) +   # reliable high-danger shooters
  annotate("rect",
           xmin = -Inf, xmax = 0,
           ymin = -Inf, ymax = hline,
           fill = "red", alpha = 0.06) +     # consistent underperformers
  annotate("rect",
           xmin = 0,    xmax = Inf,
           ymin = hline, ymax = Inf,
           fill = "yellow", alpha = 0.06) +  # high skill but inconsistent
  annotate("rect",
           xmin = -Inf, xmax = 0,
           ymin = hline, ymax = Inf,
           fill = "gray", alpha = 0.06) +    # below average + inconsistent

  # size points by CI width big==more uncertain 
  geom_point(aes(size = ci_width), alpha = 0.65) +
  scale_size_continuous(
    name   = "Uncertainty\n(95% CI width)",
    range  = c(1, 5)
  ) +
  
  geom_vline(xintercept = vline, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = hline, linetype = "dashed", color = "gray40") +
  
  # quadrant annotation labels
  annotate("text", x = Inf,  y = -Inf,  hjust = 2.1, vjust = -0.5,
           label = "Reliable High-Danger", size = 2.8, color = "darkgreen") +
  annotate("text", x = -Inf, y = -Inf,  hjust = -0.1, vjust = -0.5,
           label = "Consistent underperformers", size = 2.8, color = "darkred") +
  annotate("text", x = Inf,  y = Inf,   hjust = 2.1, vjust = 1.5,
           label = "Skilled but inconsistent", size = 2.8, color = "goldenrod4") +
  annotate("text", x = -Inf, y = Inf,   hjust = -0.1, vjust = 1.5,
           label = "Below Avg. + inconsistent", size = 2.8, color = "gray40") +

  ggrepel::geom_text_repel(
    data = player_summary %>%
      filter(abs(skill_median) > quantile(abs(skill_median), 0.90) |
             drift_median      > quantile(drift_median, 0.90)),
    aes(label = player_name),
    size = 3, max.overlaps = 15
  ) +
  
  labs(
    title    = "xG Surface: Average Deviation vs. Temporal Stability",
    x        = "Average Deviation from Population xG Surface (log-odds)",
    y        = "xG Surface Drift Across Time Blocks",
    color    = "Position"
  ) +
  theme_bw() +
  theme(legend.position = "right")
```

Forecast players next block xG
```{r}
player_blocks %>% 
  select(player_name) %>% 
  distinct() %>% 
  arrange(player_name) %>%
  print(n = 50)
```

```{r}
forecast_xg <- function(player_name_input, n_future_blocks = 3) {
  
  player_data <- player_blocks %>%
    filter(player_name == player_name_input)
  
  #safety check
  if (nrow(player_data) == 0) {
    stop(paste("Player not found:", player_name_input, 
               "\nCheck player_blocks %>% distinct(player_name)"))
  }
  
  last_block  <- max(player_data$time_block)
  future_data <- tibble(
    player_id   = unique(player_data$player_id),
    player_name = player_name_input,
    time_block  = seq(last_block + 1, last_block + n_future_blocks),
    shots       = round(mean(player_data$shots)),
    mean_x      = mean(player_data$mean_x),
    mean_y      = mean(player_data$mean_y)
  )
  
  preds <- posterior_predict(
    fit_dynamicmod,          
    newdata          = future_data,
    allow_new_levels = FALSE
  )
  
  future_data %>%
    mutate(
      pred_goals_median = apply(preds, 2, median),
      pred_goals_lower  = apply(preds, 2, quantile, 0.025),
      pred_goals_upper  = apply(preds, 2, quantile, 0.975),
      pred_xg           = pred_goals_median / shots
    )
}
```

```{r}
# name output
forecast_xg("Fagemo, Samuel", n_future_blocks = 3)
forecast_xg("Gaudette, Adam", n_future_blocks = 3)

forecast_xg("Gawdin, Glenn", n_future_blocks = 3)
forecast_xg("Vrana, Jakub", n_future_blocks = 3)
forecast_xg("Smejkal, Jiri", n_future_blocks = 3)
```

```{r}
# Combine most unquie players forecasts into one table
bind_rows(
  forecast_xg("Fagemo, Samuel",  n_future_blocks = 3),
  forecast_xg("Gaudette, Adam",  n_future_blocks = 3),
  forecast_xg("Gawdin, Glenn",   n_future_blocks = 3),
  forecast_xg("Vrana, Jakub",    n_future_blocks = 3),
  forecast_xg("Smejkal, Jiri",   n_future_blocks = 3)
) %>%
  select(player_name, time_block, shots,
         pred_goals_median, pred_goals_lower, 
         pred_goals_upper, pred_xg) %>%
  mutate(pred_xg = scales::percent(pred_xg, accuracy = 0.1)) %>%
  kable(
    digits  = 2,
    caption = "Posterior Predicted Goals and Implied xG — Next 3 blocks",
    col.names = c("Player", "Block", "Shots",
                  "Goals (median)", "Goals (lower)", 
                  "Goals (upper)", "xG")
  )
```



##3. How does each puck action change the probability of generating a dangerous offensive outcome (e.g., shot or slot shot) within the next 10 seconds, given spatial and game context?

```{r}
# Check event types
events %>% count(event_type) %>% arrange(desc(n))

# Check flags + detail columns
events %>% count(flags)
events %>% count(detail)
```

Define what a dangerous outcome is + build the outcome window.
```{r}
#sort all events chronologically within each game
events_seq <- events %>%
  arrange(game_id, period, period_time) %>%
  group_by(game_id) %>%
  mutate(
    # absolute time in seconds within game
    abs_time = (as.numeric(period) - 1) * 1200 + period_time,
    event_idx = row_number()
  ) %>%
  ungroup()

# For each event, look ahead X seconds and flag if a shot/goal occurs
#binary outcome model will predict
WINDOW_SECONDS <- 10  #try 6, 10, 15

events_labeled <- events_seq %>%
  group_by(game_id) %>%
  mutate(
    # for each event, find the next shot/goal within the window
    dangerous_next = map_lgl(seq_along(abs_time), function(i) {
      future <- abs_time[(i+1):min(i+20, n())]
      future_types <- event_type[(i+1):min(i+20, n())]
      future_team  <- team[(i+1):min(i+20, n())]
      
      # same team must generate the shot within the window
      any(
        future_types %in% c("shot", "goal") &
        future_team == team[i] &
        (future - abs_time[i]) <= WINDOW_SECONDS &
        (future - abs_time[i]) > 0,
        na.rm = TRUE
      )
    })
  ) %>%
  ungroup()

```

```{r}
#Feature engineering
model_data <- events_labeled %>%
  filter(
    !is.na(x_adj),
    !is.na(y_adj),
    !is.na(event_type),
    !is.na(player_id),
    # exclude stoppages and faceoffs
    !event_type %in% c("faceoff", "stoppage", "period_end")
  ) %>%
  mutate(
    outcome_bin  = as.integer(dangerous_next),
    is_successful = as.integer(outcome == "successful"),
    
    # normalize to offensive zone
    x_offense    = abs(x_adj),
    y_offense    = ifelse(x_adj > 0, -y_adj, y_adj),
    x_scaled     = (x_offense - mean(x_offense, na.rm = TRUE)) / 
                    sd(x_offense, na.rm = TRUE),
    y_scaled     = (y_offense - mean(y_offense, na.rm = TRUE)) / 
                    sd(y_offense, na.rm = TRUE),
    #flag into binary features  
    is_timer= str_detect(flags, "1timer"),
    is_seam= str_detect(flags, "seam"),
    is_lowhigh= str_detect(flags, "lowhigh"),
    
    # Rush situation
is_odd_man    = str_detect(detail, "1on0|2on1|3on1|3on2"),
is_even_rush  = str_detect(detail, "1on1|2on2|3on3"),

# Shot quality
is_slot_shot  = str_detect(detail, "slot") & event_type == "shot",
is_deflection = str_detect(detail, "deflection"),

# Pass danger
is_slot_pass  = detail == "slot" & event_type == "pass",
is_rush_pass  = str_detect(detail, "rush") & event_type == "pass",
is_zone_entry = str_detect(detail, "ozentry"),
    
    
    # clean event type as factor for fixed effects
    action_type  = fct_lump_n(as.factor(event_type), n = 8) %>%  # keep top 8 types
      relevel(ref = "pass"), 
      
  ) %>%
  filter(!is.na(outcome_bin))
```

```{r}
#outcome rate by action type
model_data %>%
  group_by(action_type) %>%
  summarise(
    n            = n(),
    danger_rate  = mean(outcome_bin)
  ) %>%
  arrange(desc(danger_rate))
```
*Baseline danger rate is ~26% for passes which is set as the reference level, bc passes is the most common action along with being in the middle of the danger spectrum.

```{r}
#Model formula
part3_formula <- bf(
  outcome_bin ~ action_type +
                x_scaled + y_scaled +
                I(x_scaled^2) + I(y_scaled^2) +
                is_odd_man +      # rush situation
                is_even_rush +
                is_slot_shot +    # shot from slot
                is_slot_pass +    # pass to slot
                is_rush_pass +    # pass off rush
                is_zone_entry +   # zone entry pass/reception
                is_timer +
                is_seam +
                is_successful +
                period +
                (1 | player_id)
)
```
# Fixed effects: action type + spatial location + period
# Random intercept: player_id (some players consistently generate danger)
# Random slope: action_type per player 

```{r}
# Check baseline danger rate to set intercept prior
mean(model_data$outcome_bin)
```

```{r}
#Priors
# Baseline: if ~20-30% of actions lead to danger within 10s,log-odds, log(0.25/0.75)=  -1.1
part3_priors <- c(
  # baseline probability of dangerous outcome
  prior(normal(-1, 1),   class = Intercept),
  
  # wider than parts 1/2 because action effects could be substantial
  prior(normal(0, 1.5),  class = b),
  # odd man rushes and 1-timers should have stronger effects
  # slightly wider prior to allow large positive coefficients
  prior(normal(0, 1.5),  class = b, coef = "is_timerTRUE"),
  prior(normal(0, 1.5),  class = b, coef = "is_seamTRUE"),
  prior(normal(0, 1.5), class = b, coef = "is_successful"),
  prior(normal(0, 2), class = b, coef = "is_odd_manTRUE"),
prior(normal(0, 2), class = b, coef = "is_slot_shotTRUE"),
prior(normal(0, 2), class = b, coef = "is_slot_passTRUE"),
 
  # player random intercept
  prior(exponential(2),  class = sd)
)
```

```{r}
# CODE IS SOOOO SLOW
#Randomly sample 20% of rows aka still 200k+ observations
set.seed(42)
model_data_small <- model_data %>%
  slice_sample(prop = 0.20)

# Check outcome rate is preserved
mean(model_data_small$outcome_bin)  # should still be about 0.22
```
```{r}
if (!file.exists("fit_3.rds")) {
  fit_3 <- brm(
    formula  = part3_formula,
    data     = model_data_small,  # subsampled data
    family   = bernoulli(link = "logit"),
    prior    = part3_priors,
    chains   = 2,
    iter     = 1000,
    warmup   = 500,
    cores    = 4,
    seed     = 42,
    control  = list(adapt_delta = 0.95, max_treedepth = 12)
  )
  saveRDS(fit_3, "fit_3.rds")
} else {
  fit_3 <- readRDS("fit_3.rds")
}
```


```{r}
#Fit model
if (!file.exists("modfit_3.rds")) {
  modfit_3 <- brm(
    formula  = part3_formula,
    data     = model_data_small,
    family   = bernoulli(link = "logit"),
    prior    = part3_priors,
    chains   = 4,
    iter     = 2000,
    warmup   = 1000,
    cores    = 4,
    seed     = 42,
    control  = list(adapt_delta = 0.95,
    max_treedepth = 12)
  )
  saveRDS(modfit_3, "modfit_3.rds")
} else {
  modfit_3 <- readRDS("modfit_3.rds")
}
```

```{r}
#Diagnostics
summary(fit_3)   # Rhat 
plot(fit_3)
pp_check(fit_3, ndraws = 100)
```

```{r}
# Extract posterior draws for action type coefficients
action_effects <- as_draws_df(fit_3) %>%
  select(starts_with("b_action_type")) %>%
  pivot_longer(everything(), names_to = "action", values_to = "log_odds") %>%
  mutate(
    action = str_remove(action, "b_action_type"),
    # convert log-odds shift to probability scale relative to baseline
    prob_shift = plogis(log_odds) - plogis(0)
  )

#Plot
action_effects %>%
  ggplot(aes(x = log_odds, y = reorder(action, log_odds))) +
  stat_halfeye(.width = c(0.66, 0.95)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#800000") +
  labs(
    title    = "Effect of Puck Action on P(dangerous outcome within 10s)",
    x        = "Log-odds Coefficient",
    y        = NULL
  )
```

Spatial heatmap
```{r}
# Extract exact levels the model was trained on
model_action_levels <- fit_3$data$action_type %>% levels()
model_period_levels <- fit_3$data$period %>% levels()

# Check 
model_action_levels
model_period_levels

grid_p3 <- expand_grid(
  x_offense = seq(0, 100, length.out = 60),
  y_offense = seq(-30, 30, length.out = 60)
) %>%
  mutate(
    x_scaled      = (x_offense - mean(model_data_small$x_offense)) / 
                     sd(model_data_small$x_offense),
    y_scaled      = (y_offense - mean(model_data_small$y_offense)) / 
                     sd(model_data_small$y_offense),
    
    # use levels directly from fitted model
    action_type   = factor("pass", levels = model_action_levels),
    period        = 2L,
    player_id     = NA,
    
    is_odd_man    = factor(FALSE, levels = c(FALSE, TRUE)),
    is_even_rush  = factor(FALSE, levels = c(FALSE, TRUE)),
    is_slot_shot  = factor(FALSE, levels = c(FALSE, TRUE)),
    is_slot_pass  = factor(FALSE, levels = c(FALSE, TRUE)),
    is_rush_pass  = factor(FALSE, levels = c(FALSE, TRUE)),
    is_zone_entry = factor(FALSE, levels = c(FALSE, TRUE)),
    is_timer      = factor(FALSE, levels = c(FALSE, TRUE)),
    is_seam       = factor(FALSE, levels = c(FALSE, TRUE)),
    is_successful = 1L
  )

grid_p3$danger_prob <- posterior_epred(
  fit_3,
  newdata          = grid_p3,
  re_formula       = NA,
  allow_new_levels = TRUE
) %>% colMeans()

ggplot(grid_p3, aes(x_offense, y_offense, fill = danger_prob)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", labels = scales::percent) +
  labs(
    title    = "P(dangerous outcome within 10s) by location",
    fill     = "P(danger)",
    x        = "x (offense)",
    y        = "y (offense)"
  )
#Baseline: successful pass
```


Per-action danger probability summary table.
```{r}
# Marginal predicted danger probability per action type
# Fix action_summary to match model levels
action_summary <- model_data_small %>%
  mutate(action_type = factor(action_type, levels = model_action_levels)) %>%
  filter(!is.na(action_type)) %>%
  group_by(action_type) %>%
  summarise(
    n             = n(),
    observed_rate = mean(outcome_bin),
    .groups       = "drop"
  )

#posterior predicted probability for each action type at mean location
newdata_actions <- model_data_small %>%
  mutate(action_type = factor(action_type, levels = model_action_levels)) %>%
  filter(!is.na(action_type)) %>%
  group_by(action_type) %>%
  summarise(
    x_scaled      = mean(x_scaled),
    y_scaled      = mean(y_scaled),
    period        = 2L,
    player_id     = NA,
    is_odd_man    = factor(FALSE, levels = c(FALSE, TRUE)),
    is_even_rush  = factor(FALSE, levels = c(FALSE, TRUE)),
    is_slot_shot  = factor(FALSE, levels = c(FALSE, TRUE)),
    is_slot_pass  = factor(FALSE, levels = c(FALSE, TRUE)),
    is_rush_pass  = factor(FALSE, levels = c(FALSE, TRUE)),
    is_zone_entry = factor(FALSE, levels = c(FALSE, TRUE)),
    is_timer      = factor(FALSE, levels = c(FALSE, TRUE)),
    is_seam       = factor(FALSE, levels = c(FALSE, TRUE)),
    is_successful = 1L,
    .groups       = "drop"
  )

# Verify levels match
levels(newdata_actions$action_type)

#posterior predictions
pred_probs <- posterior_epred(
  fit_3,
  newdata          = newdata_actions,
  re_formula       = NA,
  allow_new_levels = TRUE
)

# Join preds to summary
action_summary <- action_summary %>%
  mutate(
    posterior_mean  = colMeans(pred_probs),
    posterior_lower = apply(pred_probs, 2, quantile, 0.025),
    posterior_upper = apply(pred_probs, 2, quantile, 0.975)
  ) %>%
  arrange(desc(posterior_mean))

print(action_summary)

# Plot
action_summary %>%
  ggplot(aes(x = posterior_mean, 
             y = reorder(action_type, posterior_mean),
             xmin = posterior_lower,
             xmax = posterior_upper)) +
  geom_pointrange() +
  geom_vline(xintercept = mean(model_data_small$outcome_bin),
             linetype = "dashed", color = "gray50") +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title    = "Posterior Predicted Danger Probability by Action Type",
    subtitle = "Dashed line = Overall Observed Danger Rate (21.9%)",
    x        = "P(dangerous outcome within 10s)",
    y        = NULL
  )
```

Contextual Features Effect 
```{r}
# Extract all contextual binary feature posteriors
context_effects <- as_draws_df(fit_3) %>%
  select(
    "One-Timer"      = b_is_timerTRUE,
    "Successful"     = b_is_successful,
    "Slot Pass"      = b_is_slot_passTRUE,
    "Rush Pass"      = b_is_rush_passTRUE,
    "Seam Pass"      = b_is_seamTRUE,
    "Zone Entry"     = b_is_zone_entryTRUE,
    "Slot Shot"      = b_is_slot_shotTRUE
  ) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "log_odds") %>%
  mutate(
    prob_shift = plogis(log_odds) - plogis(0)
  )

context_effects %>%
  ggplot(aes(x = log_odds, y = reorder(feature, log_odds))) +
  stat_halfeye(.width = c(0.66, 0.95), fill = "#800000") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title    = "Effect of Puck Action Context on P(danger within 10s)",
    x        = "Log-odds coefficient",
    y        = NULL
  ) +
  theme_bw()
```

Compare Same Action With vs Without Key Contexts
```{r}
#comparison scenarios
scenarios <- tibble(
  scenario      = c("Pass — baseline",
                    "Pass + one-timer",
                    "Pass + slot pass",
                    "Pass + rush pass",
                    "Pass + seam",
                    "Carry — baseline",
                    "Reception — baseline"),
  action_type   = factor(c("pass","pass","pass","pass","pass","carry","reception"),
                         levels = model_action_levels),
  is_timer      = factor(c(FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,FALSE), levels=c(FALSE,TRUE)),
  is_slot_pass  = factor(c(FALSE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE), levels=c(FALSE,TRUE)),
  is_rush_pass  = factor(c(FALSE,FALSE,FALSE,TRUE,FALSE,FALSE,FALSE), levels=c(FALSE,TRUE)),
  is_seam       = factor(c(FALSE,FALSE,FALSE,FALSE,TRUE,FALSE,FALSE), levels=c(FALSE,TRUE)),
  is_odd_man    = factor(FALSE, levels=c(FALSE,TRUE)),
  is_even_rush  = factor(FALSE, levels=c(FALSE,TRUE)),
  is_slot_shot  = factor(FALSE, levels=c(FALSE,TRUE)),
  is_zone_entry = factor(FALSE, levels=c(FALSE,TRUE)),
  x_scaled      = 0,
  y_scaled      = 0,
  period        = 2L,
  player_id     = NA,
  is_successful = 1L
)

scenario_preds <- posterior_epred(
  fit_3,
  newdata          = scenarios,
  re_formula       = NA,
  allow_new_levels = TRUE
)

scenarios <- scenarios %>%
  mutate(
    pred_mean  = colMeans(scenario_preds),
    pred_lower = apply(scenario_preds, 2, quantile, 0.025),
    pred_upper = apply(scenario_preds, 2, quantile, 0.975)
  )

ggplot(scenarios,
       aes(x = pred_mean,
           y = reorder(scenario, pred_mean),
           xmin = pred_lower,
           xmax = pred_upper)) +
  geom_pointrange(size = 0.8) +
  geom_vline(xintercept = mean(model_data_small$outcome_bin),
             linetype = "dashed", color = "gray50") +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title    = "How Context Transforms Danger Probability",
    x        = "P(dangerous outcome within 10s)",
    y        = NULL
  ) +
  theme_bw()
```


Player Random Effects for Danger Generation
```{r}
player_danger <- fit_3 %>%
  spread_draws(r_player_id[player_id,]) %>%
  median_qi(.width = 0.95) %>%
  left_join(
    model_data_small %>% distinct(player_id, player_name),
    by = "player_id"
  ) %>%
  arrange(desc(r_player_id))

# Top 15 and bottom 15
player_danger %>%
  slice(c(1:15, (n()-14):n())) %>%
  ggplot(aes(x = r_player_id,
             y = reorder(player_name, r_player_id),
             xmin = .lower,
             xmax = .upper)) +
  geom_pointrange() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title    = "Player Random Effects: Danger Generation Beyond Action Context",
    x        = "Random Intercept (log-odds)",
    y        = NULL
  ) +
  theme_bw()
```
